\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology

### 2022-02-04 jsdse/Manuscript.tex
beefing it up

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

@@

%\bibliographystyle{natbib}

@@

\if0\blind
{
  \title{\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge "Sustaining the Open Tree of Life", NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

@@

Reproducibility --the extent to which consistent results are obtained when a scientific
research experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement
of science, as it constitutes a minimum standard that allows understanding scientific products
(methods, analysis, results),
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
(\cite{king1995replication, peng2011reproducible, powers2019open}).
In the natural sciences, rates of reproducibility are low (\cite{ioannidis2005most, prinz2011believe}),
prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}).
%%% The scientific community has united to incentivize cultural changes that will improve
%%% reproducibility rates long term, such as transparency, availability, and workflow
%%% automatization, to name a few (\cite{peng2015reproducibility}).
In response, the scientific community has been generating standards and incentivizing
cultural changes in an effort to improve reproducibility rates long term
(\cite{peng2015reproducibility, wilkinson2016fair}).
One standard that has received much attention is scientific availability, which
 we define as the property of research objects to be acquired, copied, analyzed,
  processed and/or reused, at no financial, legal or technical cost (\cite{arnold2019turing}),
   and with no geographic, demographic or temporal barriers (\cite{fecher2014open}).
In this paper, we argue that availability can not be fully achieved without accessibility (Box 1)
We identify factors that have particularly affected accessibility in
the natural sciences, and ways researchers can address them to ensure reproducible
and automatic scientific workflows.
% full workflow automatization and reproducibility in the natural sciences

@@

\noindent\fbox{%
    \parbox{\textwidth}{%
    \textbf{Box 1. The role of accessibility to achieve full availability.}
      Availability is generally defined as the ability or potential of something
      to be used and accessed (\cite{availability2022cambridge}).
      Following this definition, accessibility is generally considered a synonym of availability.
      Yet, in practice, availability does not imply accessibility.
      One example we find really useful to practically differentiate the concepts of availability and accessibility
      is the "marshmallows in an office" story, which goes like this.
      A couple of brand new marshmallow bags remained in a common office area after
      a social gathering.
      The marshmallows were not claimed by anybody, so they were left in the common are,
      freely available to be taken or eaten by anyone in the office.
      Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
      It was not until someone decided to open the bags and placed the marshmallows on a tray,
      that people started actually eating them. They were gone in a matter of hours.
      Scientific products are not sweet as marshmallows, but they are a coveted resource.
      While sharing, describing and documenting makes scientific products available,
      it does not necessarily makes them accessible. For example, documentation of an
      available scientific workflow might be written using expert language that is too foreign
      for a general audience to successfully reproduce the workflow. While some individuals might
      be able to invest the time to learn the expert language to successfully reproduce
      the workflow, this will not be the case for the majority, even if it is
      something that would be useful long term, the short term time investment
      is often too high for individuals to engage with it, making the workflow and its results
      effectively inaccessible and hence unavailable to the majority of the community.
    }%
}
\bigskip

To elaborate on our thesis, we use an example in evolutionary biology that focuses
on understanding the shared ancestry of organisms through phylogenetic trees, which provide the
basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
As such, improving reproducibility rates
in phylogenetic research is relevant for many aspects of biological research.
The Open Tree of Life project (OpenTree) has developed a computational platform that provides
availability of results from phylogenetic research, by creating a standardized database of
phylogenetic data that is used to synthesize a single phylogenetic tree encompassing
all life (\cite{opentreeoflife2019synth}).
The OpenTree database Phylesystem (\cite{}) is open access and available programmatically through its many
Application Programming Interface (API) services (\cite{opentreeAPIs}).
Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
to make them available to a wider programming audience.
The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research (\cite{sanchez2019datelife}) and education
(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).

The OpenTree project demonstrates that efforts to increase availability for reproducibility
in phylogenetics have also increased baseline computational knowledge and skills
required in the field.
"Computers are playing an increasingly important role in many scientific disciplines" (\cite{piccolo2016tools}).
Baseline computational skill requirements are increasing across all natural sciences.
Yet, computing is not traditionally a core skill taught to biologists and naturalists
(\cite{sayres2018bioinformatics, wright2019the, williams2019barriers}).
Still, many students are now being trained in R programming language as a statistics and data analysis environment.
% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
In order for reproducible computational tools to be adopted for research, they
 need to be much more accessible for present and future researchers.
Data and code availability is a core requirement for reproducibility research
 (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
However, the utility of data resources is limited by the technical challenges of accessing the data.
In order to motivate reproducible research, that gap needs to be bridged.
In this work we focus on improving accessibility of code examples and documentation.
In particular, we identify the necessity for documentation that is written down
 using language that is common to the target audience to facilitate examination,
  application, and adoption of code by the wider audience.

We present a set of principles to generate documentation that improves accessibility
 of code and documentation. We applied these principles to a series of tutorials
  and vignettes developed for the OpenTree project.

@@

Good primary documentation for code describes general usage of individual functions,
 the components and variables a function can take, and it should be accompanied with
  function usage examples on how to apply it.
As opposed to code, primary documentation is written in natural language (i.e.,
 any known human language, e.g., English, Spanish, Chinese) and usually makes use
  of highly specialized computational jargon (computationally specific concepts,
   words, and phrases) as well as formal language, which often slows down or even
    obstructs examination, application, and adoption of code by external individuals.
Because primary documentation is considered a professional document, acceptance
 of the research by the scientific community could diminish if a more informal
  language is used.
Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
 additional cases of individual function usage, and showcase function associations
  that work for specific analysis workflows in more detail.
As secondary documentation has become more common practice, we identify and propose
a series of principles to support improved accessibility for secondary documentation.

@@

%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
Programming languages are abstract and usually hard to remember without
an obvious link to an application (\cite{knuth1984literate}).
A story or narrative that links pieces of code together can greatly support users/students
to remember the code and its usage.
The "literate programming" paradigm (\cite(knuth1984literate))
integrates narratives with computer code in the same document, to support users
in following the code usage, improving reproducibility of scientific analyses (\cite(piccolo2016tools)).

We examined available primary documentation for the OpenTree API R wrapper (the package `rotl`),
and designed a workflow that visits as many functions as possible, and demonstrate
uses commonly requested by OpenTree users that are not demonstrated elsewhere.
By framing the function workflow using highly requested uses, the documentation acquires a
narrative arc that is easier to follow and remember by users. This facilitates the translation
of functions to specific use cases in biology.

For the tutorial demonstrated here, we used the commonly requested use case of obtaining
a phylogenetic tree for all lineages within a specific taxonomic rank.

@@

Primary documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
user adoption of reproducible workflows: demonstrate examples that do not work
as expected and exemplify ways to address them (Figure \ref{fig:first}).
We identify inputs that would give
a wide range of warnings and errors, focusing on demonstrating these cases. This
helps users to not be afraid of errors and warnings, but instead to use them to
their advantage.
We also identify effects of warnings and errors downstream of the workflow.

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
On our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and erros, and
2) it allows users/students to understand with more depth what the function is doing.

We designed ways to access the different elements of the outputs.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as "emojis", to make the language more familiar to a
broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials with an audience on workshops as well as individual user.
We choose examples that are charismatic for the audience.
For example, when we presented the tutorial for a team specialized in Amphibians,
we tailored the examples using frogs and their allies.

@@

We published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website (\cite{RopentreeTutorials, RopentreeTutorialsWebsite}).
The tutorial is available for the users to go back to it any time they need it,
and to be passed on to other users (Figure \ref{fig:second}).

@@

Ultimately, the improvement of reproducibility rates in science long term depends
on our ability to intentionally integrate the subject of reproducibility in the
the curriculum for college students and aspiring researchers to develop the skills
necessary to create reproducible scientific workflows and materials.

Some universities have been incorporating the subject in their classes (see University
 of Washington and The National Institute for General Medical Sciences).
The focus has been on students to develop skills to document their work.
The principles identified and outlined here can be used to set learning goals and
 outcomes on new reproducibility syllabus.

@@

\bibliographystyle{agsm}

>>>>

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

@@

%\bibliographystyle{natbib}
%\bibliographystyle{bmc-mathphys}
\bibliographystyle{agsm}

@@

\if0\blind
{
  \title{\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge ``Sustaining the Open Tree of Life``, NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

@@

Reproducibility --the extent to which consistent results are obtained when a scientific
research experiment is repeated \citep{repdef2021}-- is a key aspect for the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.),
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a reproducibility crisis in the field \citep{baker2016reproducibility}.
%%% The scientific community has united to incentivize cultural changes that will improve
%%% reproducibility rates long term, such as transparency, availability, and workflow
%%% automatization, to name a few \citep{peng2015reproducibility}.
In response, the scientific community has been developing new principles and standards to incentivize
cultural changes in an effort to improve long term reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair}.
A standard for reproducibility that has received much attention is scientific availability, which
we define as the property of research products to be acquired, copied, analyzed,
processed and/or reused, at no financial, legal or technical cost \citep{arnold2019turing},
and with no geographic, demographic or temporal barriers for the population \citep{fecher2014open}.
In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).
We identify factors that have particularly affected accessibility in
the natural sciences, and ways scientific developers can address them to ensure reproducible
and automatic scientific workflows. This principles can be generalized and integrated
into the reproducibility curriculum to hopefully provide present learners and future researchers with
the tools for successful scientific reproducibility.
% full workflow automatization and reproducibility in the natural sciences

@@

\noindent\fbox{%
    \parbox{\textwidth}{%
    \textbf{Box 1. The role of accessibility to achieve full availability.}
      Availability is generally defined as the ability or potential of something
      to be used and accessed \citep{availability2022cambridge}.
      Following this definition, accessibility is generally considered a synonym of availability.
      Yet, in practice, availability does not imply accessibility.
      One example we find really useful to practically differentiate the concepts of availability and accessibility
      is the ``marshmallows in an office`` story, which goes like this.
      A couple of brand new marshmallow bags remained in a common office area after
      a social gathering.
      The marshmallows were not claimed by anybody, so they were left in the common are,
      freely available to be taken or eaten by anyone in the office.
      Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
      It was not until someone decided to open the bags and placed the marshmallows on a tray,
      that people started actually eating them. They were gone in a matter of hours.
      Scientific products are not sweet as marshmallows, but they are a coveted resource.
      While sharing, describing and documenting makes scientific products available,
      it does not necessarily makes them accessible. For example, documentation of an
      available scientific workflow might be written using expert language that is too foreign
      for a general audience to successfully reproduce the workflow. While some individuals might
      be able to invest the time to learn the expert language to successfully reproduce
      the workflow, this will not be the case for the majority, even if it is
      something that would be useful long term, the short term time investment
      is often too high for individuals to engage with it, making the workflow and its results
      effectively inaccessible and hence unavailable to the majority of the community.
    }%
}
\bigskip

To elaborate on our thesis, we use a widely used scientific workflow in phylogenetics,
a research field within evolutionary biology that focuses
on understanding the shared ancestry of living organisms through the reconstruction of
phylogenetic trees from biological data. Phylogenies provide the
basis to study and understand all biological processes in an evolutionary context
\citep{dobzhansky1973nothing}.
Given the importance of phylogenies, improving reproducibility rates
in phylogenetic research is increasingly relevant for many aspects of the natural sciences.
The Open Tree of Life project (OpenTree) has developed a computational platform that provides
availability of results from phylogenetic research, by creating a standardized database of
phylogenetic data that is used to synthesize a single phylogenetic tree encompassing
all life \citep{opentreeoflife2019synth}.
The OpenTree database Phylesystem \citep{mctavish2015phylesystem} is programmatically
available free of cost to users through OpenTree's Application Programming Interface
(API) services \citep{opentreeAPIs}.
Additionally, software packages written with open source, free of cost programming
languages commonly used in scientific research such as R and Python
\citep{baker2017scientific} have been developed as wrappers that have increased the availability
of OpenTree's API services, by making the OpenTree API functionalities more accessible
to a wider user audience \citep{michonneau2016rotl, mctavish2021opentree}.
Yet, the R and Python OpenTree API wrapper software packages have been mainly used
by computer-literate individuals
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research \citep{sanchez2019datelife, sanchez2021physcraper}
and education \citep{nguyen2020phylotastic, phylotasticedtools, galacticedtools}.
In the 5 years since its release, the R package wrapper for OpenTree \texttt{rotl}
has been cited by more than 190 biological research papers addressing various
topics \citep{googlerotl}.
Comparatively, other R packages for phylogenetics, such as \texttt{GGTREE}
and \texttt{mixOmics}, which were released one year after \texttt{rotl} have been
cited over a thousand times, which might indicate that \texttt{rotl} has not been
adopted as widely.
%% We need a proper statistical distribution of citation numbers per year, to see if rotl is an outlier or not.

While learners in the natural sciences have been engaging independently with R
and Python programming language
as they represent two of the most widely used programming languages in the sciences
% for their fast and seamless statistics and data analysis environment
\citep{baker2017scientific}, computer programming is not traditionally
a core skill formally taught to biologists and naturalists
\citep{sayres2018bioinformatics, wright2019the, williams2019barriers},
However, the OpenTree project demonstrates that efforts to improve availability and reproducibility
in phylogenetics have increased required baseline computational knowledge and skills
in the field.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
baseline computational skill requirements are also increasing across all natural sciences.
% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
% In order for reproducible computational tools to be adopted for research, they
% need to be readily accessible for present and future researchers.
Thus, efforts to increase reproducibility rates in the natural sciences must consider
the particularities of scientific workflows that rely on usage of programming languages, such as
availability and accessibility of data and code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
% However, the utility of data resources is limited by the technical challenges of accessing the data.
% In order to motivate reproducible research, that gap needs to be bridged.
% In this work we focus on improving accessibility of code examples and documentation.

We chose a phylogenetics workflow that relies on data and code from OpenTree to
identify the specific barriers to accessibility and reproducibility it presents.
Then, we design ways to overcome or diminish these barriers and apply them to
a series of tutorials and vignettes developed for the OpenTree project.
Finally, we generalize our finding to suggest a set of
principles that can be used as guide to develop code and documentation materials
with increased accessibility that can improve availability and contribute to reproducibility.
Notably, these principles can be incorporated as learning goals into any syllabus for a
course or workshop on best practices for scientific reproducibility.
% In particular, we identify the necessity for documentation that is written down
% using language that is common to the target audience to facilitate examination,
% application, and adoption of code by the wider audience.

@@

Good primary documentation for code is thorough. It describes general usage of individual functions,
the components and variables a function can take, and it should be accompanied with
function usage examples on how to apply it \citep{karimzadeh2018top}.
As opposed to code, primary documentation is written in natural language (i.e.,
any known human language, e.g., English, Spanish, Chinese).
Primary documentation is viewed as a key element for success of a piece of code
\citep{karimzadeh2018top},
which might be why it is also usually written using highly specialized computational jargon (i.e., computationally specific concepts,
words, and phrases) as well as formal scientific language.
While this might be important for formal acceptance of the code by the scientific
and academic community, it often slows down or even
obstructs examination, application, and adoption of code by the general audience \citep{ball2017its}.

Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
additional cases of individual function usage, and showcase function associations
that work for specific analysis workflows in more detail.
As secondary documentation has become more common practice and is more flexible
in its form and content, it constitutes an ideal canvas to develop, implement and
test principles that can overcome current barriers to code accessibility.
The tutorials are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.

@@

%% I think I have three different points here:
%% internalizing knowledge is made easier with active learning practices: learn by doing!
%% literate programming allows learning by doing AND at the same time provides a story that connects the acquired knowledge to a real worl problem.
%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
Pedagogical research shows that active learning practices are one the most effective
ways to take on abstract subjects \citep{freeman2014active}.
Programming computer languages are quite abstract and learning them can be greatly
enhanced by applying an active learning strategy such as a linking its usage to
a ``real world'' or ``human'' application \citep{felder2009active}.

A story-like narrative that links pieces of code together and invites learners
to try the code can greatly support learners to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following, remembering and understanding
the code usage \citep{piccolo2016tools}.
Documents developed with ``literate programming'' can be made more accessible by
choosing narratives that are relatable to a more general audience.

We examined available primary documentation for the package \texttt{rotl},
and designed a narrative that required the usage of as many functions as possible.
We demonstrate code applications that are commonly requested by OpenTree users,
but that are not demonstrated in the R package primary documentation.
By framing the function workflow using highly requested uses, the documentation acquires a
narrative arc that is easier to follow and remember by users. This can also facilitate the application
of code to other use cases in biology of interest for the learners.

%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested use case of obtaining
%% a phylogenetic tree for all lineages within a specific taxonomic rank.

@@

A practice that has become more and more widespread in programming languages pedagogical practices
is the use of typos and mistakes to normalize them for learners and show them how
to solve them when they are outside the classroom \citep{shannon2015live}.
Yet, this is rarely done for written pedagogical materials.
Primary documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
\citep{gaspar2007restoring}.
We demonstrate examples that do not work
as expected and exemplify ways to address them (Figure \ref{fig:first}).

We identify inputs that would give
a wide range of warnings and errors, focusing on demonstrating these cases. This
helps users to not be afraid of errors and warnings, but instead to use them to
their advantage.
We also identify effects of warnings and errors downstream of the workflow.

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
On our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and errors, and
2) it allows users/students to understand with more depth what the function is doing.

% We designed ways to access the different elements of the outputs.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials with an audience on workshops as well as individual user.
We choose examples that are charismatic for the audience.
For example, when we presented the tutorial for a team specialized in Amphibians,
we tailored the examples using frogs and their allies.

@@

We published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
The tutorial is available for the users to go back to it any time they need it,
and to be passed on to other users (Figure \ref{fig:second}).

@@

Ultimately, the long term improvement of reproducibility rates in science will depend
on our ability to intentionally integrate the subject of reproducibility into the
undergraduate curriculum, so college learners and future researchers have the
basis to develop the fundamental skills needed to successfully create reproducible
scientific workflows and materials.

Some universities have been incorporating the subject in their classes (see
\cite{uwlibraries2022, nigms2022}).
The focus of this resources has been for students to develop skills to document their work.
The principles identified and outlined here can be used to set learning goals and
outcomes on new reproducibility syllabus.

@@
